Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-16 demystify.tex
updates

@@

\begin{itemize}
\item Each applicant is assigned to \emph{about 3 reviewers} who all seem to have expertise in the applicant's research topics. This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite in depth. This is different from PhD applications where adcom reviewers (\autoref{sec:adcom}) might not be in your field.

Note that if sometimes an original reviwer is no longer available then someone who is not in the field might review your application.  Thus in general you want to make your application accessible to a general audience in CS.

\item Reviewers attempt to compare applicants \emph{within their cohort level}.  This means that we do not compare an undergrad, who is still applying to grad school, to a 2nd year PhD student who already has an adviser and published multiple papers. This ensures that each applicant is evaluated fairly based on their level.

\item Unlike CS PhD applications, which focus on research potentials in which concrete evidence like papers, research experience, and LoRs are important, NSF GRFP also focuses on \emph{potential} but also on broader impacts, e.g., how you can help society, how you can mentor others, etc.

@@

\subsection{Intellectual Merit and Broader Impacts}
Your will be evaluated based on two criteria: Intellectual Merit (IM) and Broader Impacts (BI).

\paragraph{Intellectual merit} is based on your research potential, e.g., your research experience, your potential to do research, and how you can contribute to the field.

Note that unlike a normal project or research proposal, in which the evaluation is based on the project, in the GRFP, the evaluation is \emph{not} about the research plan. We are repeatedly reminded to not focus too much on the research plan as many applicants are still applying for PhD programs and might not even get into the field or work with a prof. that fit their research plan.  Instead, the evaluation is based on your \emph{potential} of being a good researcher using various criteria, e.g., your research experience, what others think of you (LoRs), research plan, etc. This also means that even if you have very little research experience, you can still get the fellowship if you can convince the reviewers that you have the potential to be a good researcher.  Vice versa, if you have a lot of research experience and in very top places but cannot convince the reviewers that you have the potential to be a good researcher, you will not get the fellowship.

\paragraph{Broader impacts} based on how you and your research can benefit society.  Many students just mention how their research is cross-discipline and thus has broader impacts (e.g., improve software quality and thus help society).  This is too common and not enough.  You also want to talk about how you, as a grad student with your research, can help society, e.g., through outreach, mentoring, diversity, etc. As mentioned below, you will want to be specific and concrete, e.g., ``I will work with highschool students and underrepresented students'' is not as convincing as ``I have had experience in mentoring highschool students X and will continue to do so through Y,Z ...''.

@@

>>>>

The National Science Foundation (NSF) Graduate Research Fellowship Program (GRFP) is a prestigious fellowship for US citizens, nationals, and permanent residents who are in their first or second year of graduate study. The fellowship provides three years of financial support, whih includes a stipend and tuition. The GRFP is highly competitive, with many applicants from top schools, with strong LoR and research experience.

The GRFP website has detailed information on the application process and eligibility requirements. This document focuses on how your application is evaluated and how you can improve your chances of success.

@@

\begin{itemize}
\item Each applicant is assigned to \emph{about 3 reviewers} who have expertise in the applicant's research topics. This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite in depth. This is different from PhD applications where adcom reviewers (\autoref{sec:adcom}) might not be in your field.

Note that if sometimes an original reviwer is no longer available then someone who is not in the field might review your application.  Thus you want to make your statements accessible to a general CS audience.

\item Reviewers typically try to compare applicants \emph{within their cohort level}.  This means that we do not compare an undergrad, who is still applying to grad school, to a 2nd year PhD student who already has an adviser and published multiple papers. This ensures that each applicant is evaluated fairly based on their level.

\item Unlike CS PhD applications that focus on research potentials in which papers, research experience, and LoRs are important, NSF GRFP looks at both \emph{research potential} and \emph{broader impacts}, e.g., how you can help society, how you can mentor others, etc.  It is not expected that you have published papers or have a lot of research experience (recall that MS students can also get the fellowship). Instead, you need to convince the reviewers that you have the potential to be a good researcher (technical merits) and that you can help society (broader impacts). More in \autoref{sec:grfp-im-bi}.

@@

\subsection{Intellectual Merit and Broader Impacts}\label{sec:grfp-im-bi}

Your application will be evaluated based on two criteria: Intellectual Merit (IM) and Broader Impacts (BI).

\paragraph{Intellectual merit (IM)} IM evaluates your research potential, e.g., your research experience, your potential to do research, and how you can contribute to the field.   Unlike a normal project or research proposal, in which the evaluation is entirely based on the project, in the GRFP, the evaluation is \emph{not} entirely about the research plan. We are repeatedly reminded to not focus too much on the research plan as many applicants are still applying for PhD programs and might not even get into the field or work with a prof. that fit their research plan.

Instead, the evaluation is based on your \emph{potential} of being a good researcher using various criteria, e.g., your research experience, what others think of you (LoRs), research plan, etc. This also means that even if you have very little research experience, you can still get the fellowship if you can convince the reviewers that you have the potential to be a good researcher.  Vice versa, if you have a lot of research experience and in very top places but cannot convince the reviewers that you have the potential to be a good researcher, you will not get the fellowship.

\paragraph{Broader impacts (BM)} BM is a common and important component in NSF application. It is about how you and your research can \emph{benefit society}.  Many students just mention how their research is cross-discipline and thus has broader impacts (e.g., improve software quality and thus help society).  This is too common and not enough.  You also want to talk about how you, as a grad student with your research, can help society, e.g., through outreach, mentoring, etc. As mentioned below, you will want to be specific and concrete, e.g., \emph{``I will work with highschool students ..''} is not as convincing as \emph{``I have had experience in mentoring highschool students X and will continue to do so through Y,Z ...''}.

Writing BM is often difficult, not only for students but also new faculty.  You should ask someone who has experience writing NSF proposals to review your BM.

@@

\paragraph{Overexaggeration Lors} Many students have LoRs that overrate them, e.g., \emph{``the best student I have ever had''}, \emph{``better than most PhD students''}, etc.  This is not only not convincing but can also hurt your application.  Reviewers are often professors themselves and know that such statements are often not true.  Moreover, such statements can bias the evaluation, e.g., if a reviewer sees that a student is from a top school and has a LoR from a top professor, they might give the student a higher score than they deserve.  So you should ask your LoR writers to be honest and specific, e.g., \emph{``the student did well in my class and I think they have the potential to be a good researcher''}.
